### 2025-04-01 decision-making.qmd
20250401 - median absolute error

errors
median(errors)
mean(predictedValues)

@@

As demonstrated above, when the crowd-averaged predictions are centered on the truth, the crowd-averaged predictions are perfectly accurate and are more accurate than the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError).

Here is a code example of when the crowd-averaged predictions bracket but are not centered on the truth for a player:

@@

As demonstrated above, when the crowd-averaged predictions bracket but are not centered on the truth, the crowd-averaged predictions are more accurate than the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError).

Here is another code example of when the crowd-averaged predictions bracket but are not centered on the truth for a player:

@@

As demonstrated above, when the crowd-averaged predictions bracket but are not centered on the truth, the crowd-averaged predictions are more accurate than the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError), but they can still be less accurate than the individual forecaster who is at the 50th percentile in accuracy.

Here is a code example of when the crowd-averaged predictions do not bracket the truth for a player:

@@

As demonstrated above, when the crowd-averaged predictions do not bracket the truth, the crowd-averaged predictions are as accurate as the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError).

Here is another code example of when the crowd-averaged predictions do not bracket the truth for a player:

@@

As demonstrated above, when the crowd-averaged predictions do not bracket the truth, the crowd-averaged predictions are as accurate as the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError), but they can be less accurate than the individual forecaster who is at the 50th percentile in accuracy.

>>>>

errors
mean(predictedValues)

@@

```{r}
#| label: fig-bracketingCentered
#| fig-cap: "Example Distribution of Projected Fantasy Points that Brackets and is Centered on the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Brackets and is Centered on the Truth."
#| code-fold: true

bracketingCentered <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

maeBracketingCentered_individual <- bracketingCentered[which(row.names(bracketingCentered) == "individual"), "MAE"]
maeBracketingCentered_crowd <- bracketingCentered[which(row.names(bracketingCentered) == "crowdAveraged"), "MAE"]

mdaeBracketingCentered_individual <- bracketingCentered[which(row.names(bracketingCentered) == "individual"), "MdAE"]
meBracketingCentered_crowd <- bracketingCentered[which(row.names(bracketingCentered) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  #annotate(
  #  "segment",
  #  x = mean(predictedValues),
  #  xend = mean(predictedValues) + maeBracketingCentered_crowd,
  #  y = 0,
  #  yend = 0,
  #  linewidth = 1.5,
  #  alpha = 0.6,
  #  arrow = arrow(ends = "both")) +
  annotate(
    "text",
    x = mean(predictedValues) + (maeBracketingCentered_crowd / 2) + 1,
    y = 0.0003,
    label = paste("Mean Error: ", meBracketingCentered_crowd, sep = ""),
    hjust = 0 # left-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the crowd-averaged predictions are centered on the truth, the crowd-averaged predictions are perfectly accurate and are more accurate (MAE/MdAE = `{r} maeBracketingCentered_crowd`) than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingCentered_individual`; MdAE = `{r} mdaeBracketingCentered_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

Here is a code example of when the crowd-averaged predictions bracket but are not centered on the truth for a player, with a high bracketing rate.
The distribution of projected points versus truth is depicted in @fig-bracketingNotCentered1.

@@

```{r}
#| label: fig-bracketingNotCentered1
#| fig-cap: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| code-fold: true

bracketingNotCentered1 <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

maeBracketingNotCentered1_individual <- bracketingNotCentered1[which(row.names(bracketingNotCentered1) == "individual"), "MAE"]
maeBracketingNotCentered1_crowd <- bracketingNotCentered1[which(row.names(bracketingNotCentered1) == "crowdAveraged"), "MAE"]

mdaeBracketingNotCentered1_individual <- bracketingNotCentered1[which(row.names(bracketingNotCentered1) == "crowdAveraged"), "MdAE"]
meBracketingNotCentered1_crowd <- bracketingNotCentered1[which(row.names(bracketingNotCentered1) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  annotate(
    "segment",
    x = mean(predictedValues),
    xend = actualValue,
    y = 0,
    yend = 0,
    linewidth = 1.5,
    alpha = 0.6,
    arrow = arrow(ends = "both")
  ) +
  annotate(
    "text",
    x = actualValue + (meBracketingNotCentered1_crowd / 2),
    y = 0.0007,
    label = paste("Mean Error: ", meBracketingNotCentered1_crowd, sep = ""),
    hjust = 0.5 # center-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the crowd-averaged predictions bracket but are not centered on the truth, and the bracketing rate is high, the crowd-averaged predictions (MAE/MdAE = `{r} maeBracketingNotCentered1_crowd`) are more accurate than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingNotCentered1_individual`; MdAE = `{r} mdaeBracketingNotCentered1_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

Here is another code example of when the crowd-averaged predictions bracket but are not centered on the truth for a player, with a low bracketing rate.
The distribution of projected points versus truth is depicted in @fig-bracketingNotCentered2.

@@

```{r}
#| label: fig-bracketingNotCentered2
#| fig-cap: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| code-fold: true

bracketingNotCentered2 <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

maeBracketingNotCentered2_individual <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "individual"), "MAE"]
maeBracketingNotCentered2_crowd <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "crowdAveraged"), "MAE"]

mdaeBracketingNotCentered2_individual <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "crowdAveraged"), "MdAE"]
meBracketingNotCentered2_crowd <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  annotate(
    "segment",
    x = mean(predictedValues),
    xend = actualValue,
    y = 0,
    yend = 0,
    linewidth = 1.5,
    alpha = 0.6,
    arrow = arrow(ends = "both")
  ) +
  annotate(
    "text",
    x = actualValue + (meBracketingNotCentered2_crowd / 2),
    y = 0.0003,
    label = paste("Mean Error: ", meBracketingNotCentered2_crowd, sep = ""),
    hjust = 0.5 # center-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the crowd-averaged predictions bracket but are not centered on the truth and the bracketing rate is low, the crowd-averaged predictions (MAE/MdAE = `{r} maeBracketingNotCentered2_crowd`) are more accurate than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingNotCentered2_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError), but they can still be less accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than the individual forecaster who is at the 50th percentile in accuracy (MdAE = `{r} mdaeBracketingNotCentered2_individual`).

Here is a code example of when the crowd-averaged predictions do not bracket the truth for a player, but the distribution is still close to truth.
The distribution of projected points versus truth is depicted in @fig-noBracketing1.

@@

```{r}
#| label: fig-noBracketing1
#| fig-cap: "Example Distribution of Projected Fantasy Points that Does Not Bracket the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Does Not Bracket the Truth."
#| code-fold: true

noBracketing1 <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

maeNoBracketing1_individual <- noBracketing1[which(row.names(noBracketing1) == "individual"), "MAE"]
maeNoBracketing1_crowd <- noBracketing1[which(row.names(noBracketing1) == "crowdAveraged"), "MAE"]

mdaeNoBracketing1_individual <- noBracketing1[which(row.names(noBracketing1) == "crowdAveraged"), "MdAE"]
meNoBracketing1_crowd <- noBracketing1[which(row.names(noBracketing1) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  annotate(
    "segment",
    x = mean(predictedValues),
    xend = actualValue,
    y = 0,
    yend = 0,
    linewidth = 1.5,
    alpha = 0.6,
    arrow = arrow(ends = "both")
  ) +
  annotate(
    "text",
    x = actualValue + (meNoBracketing1_crowd / 2),
    y = 0.0003,
    label = paste("Mean Error: ", meNoBracketing1_crowd, sep = ""),
    hjust = 0.5 # center-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the crowd-averaged predictions do not bracket the truth but are still close to the truth, the crowd-averaged predictions (MAE/MdAE = `{r} maeNoBracketing1_crowd`) are as accurate as the accuracy of the typical individual forecaster (MAE = `{r} maeNoBracketing1_individual`; MdAE = `{r} mdaeNoBracketing1_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

Here is another code example of when the crowd-averaged predictions do not bracket the truth for a player and there is an outlier prediction that is far away from the truth.
The distribution of projected points versus truth is depicted in @fig-noBracketing2.

@@

```{r}
#| label: fig-noBracketing2
#| fig-cap: "Example Distribution of Projected Fantasy Points that Does Not Bracket the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Does Not Bracket the Truth."
#| code-fold: true

noBracketing2 <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

maeNoBracketing2_individual <- noBracketing2[which(row.names(noBracketing2) == "individual"), "MAE"]
maeNoBracketing2_crowd <- noBracketing2[which(row.names(noBracketing2) == "crowdAveraged"), "MAE"]

mdaeNoBracketing2_individual <- noBracketing2[which(row.names(noBracketing2) == "crowdAveraged"), "MdAE"]
meNoBracketing2_crowd <- noBracketing2[which(row.names(noBracketing2) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  annotate(
    "segment",
    x = mean(predictedValues),
    xend = actualValue,
    y = 0,
    yend = 0,
    linewidth = 1.5,
    alpha = 0.6,
    arrow = arrow(ends = "both")
  ) +
  annotate(
    "text",
    x = actualValue + (meNoBracketing2_crowd / 2),
    y = 0.0015,
    label = paste("Mean Error: ", meNoBracketing2_crowd, sep = ""),
    hjust = 0.5 # center-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the crowd-averaged predictions do not bracket the truth and there are one or more predictions that are far away from the truth, the crowd-averaged predictions (MAE/MdAE = `{r} maeNoBracketing2_crowd`) are as accurate as the accuracy of the typical individual forecaster (MAE = `{r} maeNoBracketing2_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError), but they can be less accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than the individual forecaster who is at the 50th percentile in accuracy (MdAE = `{r} mdaeNoBracketing2_individual`).

### 2025-04-01 evaluating-prediction-accuracy.qmd
20250311 - quote attribution
20250311 - add citation
20250401 - median absolute error

> "Nothing ruins fantasy like reality." – Renee Miller

@@

>>>>

> "Nothing ruins fantasy more than reality." – [Renee Miller](https://www.youtube.com/watch?v=gmpLFWs5ae0&t=1430s) [@Yahoo2024]

@@

##### Median Absolute Error (MdAE) {#sec-medianAbsoluteError}

Median absolute error (MdAE) is the median of the absolute value of differences between the predicted and observed values across individuals, that is, the median of the absolute value of errors.
Smaller MdAE values (closer to zero) reflect greater accuracy.
MdAE indicates the accuracy of the "typical" prediction (i.e., the prediction that is at the 50th percentile in terms of accuracy).
The formula for MdAE is in @eq-medianAbsoluteError:

$$
\begin{aligned}
  \text{median absolute error (MdAE)} &= \frac{\text{median}(|\text{predicted}_i - \text{observed}_i|)}{n} \\
  &= \text{median}(|e_i|)
\end{aligned}
$$ {#eq-medianAbsoluteError}
